### 2025-01-23 paper/paper.tex
Started writing the paper.

\documentclass[a4paper, 12pt]{article}

\usepackage[utf8]{inputenc} % For UTF-8 encoding
\usepackage[english]{babel} % For English language settings
% \usepackage{amsmath}        % For mathematical symbols
% \usepackage{graphicx}       % For including images
% \usepackage{hyperref}       % For hyperlinks
\usepackage[margin=1in]{geometry}       % To set margins
% \usepackage{float}          % For controlling figure placement
% \usepackage{booktabs}       % For better table formatting

\renewcommand{\baselinestretch}{1.15}

\setlength{\parindent}{0pt}
\setlength{\parskip}{0.8em}

\title{Comparison of Spatial Partitioning Strategies}
\author{Linus Kåreblom}
\date{\today}

\begin{document}

\begin{titlepage}
    \centering
    \vspace*{1cm}

    {\Huge \textbf{Comparison of Spatial Partitioning Strategies}}\\[1.5cm]

    {\large Linus Kåreblom}\\[1cm]

    \vfill

    Technical program Academic year 2024/2025\\
    \textbf{School:} Österänggymnasiet\\
    \textbf{Class:} TE3A\\
    \textbf{Supervisor:} Tony Tieu\\

    \vspace{1cm}
    {\today}
\end{titlepage}

\begin{abstract}
    This is the abstract of my paper.
\end{abstract}

\newpage

\tableofcontents
\newpage

\section{Introduction}
When I was developing a small physics engine, I encountered performance issues
during the  collision detection phase. In this phase the program figures out
which objects are close enough to be touching and overlapping. This wouldn't be
an issue in a small simulation with few objects, but I was striving for
something grand running in real time which made every millisecond count. Due to
these constraints I had to find a way to speed things up.

On my optimization journey I encountered the concept of spatial partitioning -
an elegant solution to my problem. Diving deeper I discovered the three
techniques I'll be comparing in this paper, along side the naive solution. The
basic theory of the technique is simply to group objects close to each other to
avoid checking two objects on the opposite side of the world. This drastically
decreases the number of collision checks performed each frame.

As a consequence of this constraint I had to find a way to optimize this phase.
That is what sent me down the rabbit hole of spatial partitioning as an
optimization to do query objects close to each other in space. During my
research I found out about the three strategies I'm comparing in this paper. I'm
doing this paper within the programming course.

\section{Theory}
\subsection{Spatial Partitioning}
Spatial partitioning is an optimization technique used to make programs run
faster. It's often found within physics applications. The theory behind this
technique is to reduce the number of times one object is checked against
another. Suppose we have one hundred objects in our world. The naive way to
check each one of these against the others results in a ten thousand checks.
Since checks are expensive to make, we want to perform as few of them as
possible.

\subsection{Quadtree}
A quadtree is a recursive data structure, meaning it contains pointers to each
objects of the same type as itself. It's in the family of tree data structures.
The data structure is comprised of nodes. Each node has four pointers to other
nodes, could also be referred to as branches.

\subsection{The Stack and Heap}
A program uses two different types of memory, the stack and the heap. When the
program starts a fixed amount of bytes are allocated for the program to use,
this is what is called the stack. When the program creates a variable it is
allocated on the stack, which is done by incrementing the stack pointer by the
size of the variable data type. Since a stack allocation is only a single
instruction it is incredibly fast. The heap is much larger than the stack but
slower because you have to request space within it from the operating system.
Allocating is considered to be a slow operation since there's a lot of overhead
introduced.

\subsection{Big O notation}
Big O notation is a notation system to describe the time complexity of an
algorithm. If an algorithm has the time complexity of $O(n)$ then that means
its time complexity grows linearly with the amount of inputs $n$. This could be
deceiving due to a hidden constant within the expression. If $O(n)$ would be
written as a function to describe the actual execution time of the algorithm
then it would look like this $T(n) = C \cdot n$. A $O(n^{2})$ algorithm would
be considered worse than an $O(n)$ one. This assumption is, however, not always
correct. If $n$ is small enough then the $O(n^{2})$ algorithm could run faster
due to the constant being smaller than its $O(n)$ equivalent.

\subsection{Static Grid}
The static grid is built using a simple array. Indexing into the array is an
$O(1)$ operation. The caveat with this approach is you're forced to know how
big the area you want the to partition to be before compiling the program. Both
insertion and querying only requires a simple index into the grid array which
makes this approach super fast.

\subsection{Spatial Hashing Grid}
The spatial hashing grid is built using the same technique as a hash map. The
method I used to write my spatial hashing grid was to allocate an arbitrarily
sized array. When inserting into the structure I first hash the coordinate of
the object, then run a modulo operation on it giving me a number within the
array bounds, and then it is inserted into the cell.

\end{document}

### 2025-01-23 include/grid.h
Refactored the benchmarking functions to use an interface to reduce code duplication.

@@ -20,12 +20,18 @@ struct Grid {
     Cell *cells;
 };
 
-extern Grid grid_new(Box grid_size, Vec2 cell_count);
+typedef struct GridDesc GridDesc;
+struct GridDesc {
+    Box grid_size;
+    Vec2 cell_count;
+};
+
+extern Grid* grid_new(const GridDesc* desc);
 extern void grid_free(Grid *grid);
 
 extern void grid_insert(Grid *grid, Box box);
-extern void grid_reset(Grid *grid);
+extern void grid_clear(Grid *grid);
 
-extern Vec(Box) grid_query(Grid grid, Box area);
+extern Vec(Box) grid_query(const Grid* grid, Box area);
 
-extern void grid_debug_draw(Grid grid, SDL_Renderer *renderer);
+extern void grid_debug_draw(const Grid* grid, SDL_Renderer *renderer);

### 2025-01-23 include/hashing.h

@@ -21,12 +21,18 @@ struct SpatialHash {
     Bucket *buckets;
 };
 
-extern SpatialHash spatial_hash_new(uint32_t map_capacity, Vec2 cell_size); 
+typedef struct SpatialHashDesc SpatialHashDesc;
+struct SpatialHashDesc {
+    uint32_t map_capacity;
+    Vec2 cell_size;
+};
+
+extern SpatialHash* spatial_hash_new(const SpatialHashDesc* desc);
 extern void spatial_hash_free(SpatialHash *space);
 
 extern void spatial_hash_insert(SpatialHash *space, Box box);
 extern void spatial_hash_clear(SpatialHash *space);
 
-extern Vec(Box) spatial_hash_query(SpatialHash space, Box area);
+extern Vec(Box) spatial_hash_query(const SpatialHash* space, Box area);
 
-extern void spatial_hash_debug_draw(SpatialHash space, uint32_t horizontal_count, uint32_t vertical_count, SDL_Renderer *renderer);
+extern void spatial_hash_debug_draw(const SpatialHash* space, SDL_Renderer *renderer);

### 2025-01-23 include/naive.h

@@ -1,0 +1,18 @@
+#pragma once
+
+#include "ds.h"
+#include "box.h"
+
+#include <SDL2/SDL.h>
+
+typedef struct Naive Naive;
+struct Naive {
+    Vec(Box) boxes;
+};
+
+extern Naive* naive_new(const void* desc);
+extern void naive_free(Naive* data);
+extern void naive_insert(Naive* data, Box box);
+extern void naive_clear(Naive* data);
+extern Vec(Box) naive_query(const Naive* data, Box area);
+extern void naive_debug_draw(const Naive* data, SDL_Renderer* renderer);

### 2025-01-23 include/quadtree.h

@@ -32,12 +32,19 @@ struct Quadtree {
     uint32_t max_box_count;
 };
 
-extern Quadtree quadtree_new(Box area, int max_depth, int max_box_count);
+typedef struct QuadtreeDesc QuadtreeDesc;
+struct QuadtreeDesc {
+    Box area;
+    int max_depth;
+    int max_box_count;
+};
+
+extern Quadtree* quadtree_new(const QuadtreeDesc* desc);
 extern void quadtree_free(Quadtree *quadtree);
 
 extern void quadtree_insert(Quadtree *quadtree, Box box);
 extern void quadtree_clear(Quadtree *quadtree);
 
-extern Vec(Box) quadtree_query(Quadtree quadtree, Box area);
+extern Vec(Box) quadtree_query(const Quadtree* quadtree, Box area);
 
-extern void quadtree_debug_draw(Quadtree quadtree, SDL_Renderer *renderer);
+extern void quadtree_debug_draw(const Quadtree* quadtree, SDL_Renderer *renderer);

### 2025-01-23 include/strategy_interface.h

@@ -1,0 +1,63 @@
+#pragma once
+
+#include "ds.h"
+#include "box.h"
+#include "quadtree.h"
+#include "grid.h"
+#include "hashing.h"
+#include "naive.h"
+
+#include <SDL2/SDL.h>
+
+typedef void* (*StrategyNewFunc)(const void* desc);
+typedef void (*StrategyFreeFunc)(void* data);
+typedef void (*StrategyInsertFunc)(void* data, Box box);
+typedef void (*StrategyClearFunc)(void* data);
+typedef Vec(Box) (*StrategyQueryFunc)(const void* data, Box area);
+typedef void (*StrategyDebugDrawFunc)(const void* data, SDL_Renderer* renderer);
+
+typedef struct Strategy Strategy;
+struct Strategy {
+    StrategyNewFunc new;
+    StrategyFreeFunc free;
+    StrategyInsertFunc insert;
+    StrategyClearFunc clear;
+    StrategyQueryFunc query;
+    StrategyDebugDrawFunc debug_draw;
+};
+
+static const Strategy STRATEGY_QUADTREE = {
+    .new        = (StrategyNewFunc)       quadtree_new,
+    .free       = (StrategyFreeFunc)      quadtree_free,
+    .insert     = (StrategyInsertFunc)    quadtree_insert,
+    .clear      = (StrategyClearFunc)     quadtree_clear,
+    .query      = (StrategyQueryFunc)     quadtree_query,
+    .debug_draw = (StrategyDebugDrawFunc) quadtree_debug_draw,
+};
+
+static const Strategy STRATEGY_GRID = {
+    .new        = (StrategyNewFunc)       grid_new,
+    .free       = (StrategyFreeFunc)      grid_free,
+    .insert     = (StrategyInsertFunc)    grid_insert,
+    .clear      = (StrategyClearFunc)     grid_clear,
+    .query      = (StrategyQueryFunc)     grid_query,
+    .debug_draw = (StrategyDebugDrawFunc) grid_debug_draw,
+};
+
+static const Strategy STRATEGY_SPATIAL_HASHING = {
+    .new        = (StrategyNewFunc)       spatial_hash_new,
+    .free       = (StrategyFreeFunc)      spatial_hash_free,
+    .insert     = (StrategyInsertFunc)    spatial_hash_insert,
+    .clear      = (StrategyClearFunc)     spatial_hash_clear,
[…]

### 2025-01-23 src/grid.c

@@ -5,11 +5,12 @@
 #include <math.h>
 #include <stdio.h>
 
-Grid grid_new(Box grid_size, Vec2 cell_count) {
-    Grid grid = {
-        .world_box = grid_size,
-        .cell_count = cell_count,
-        .cells = malloc(cell_count.x*cell_count.y*sizeof(Cell)),
+Grid* grid_new(const GridDesc* desc) {
+    Grid* grid = malloc(sizeof(Grid));
+    *grid = (Grid) {
+        .world_box = desc->grid_size,
+        .cell_count = desc->cell_count,
+        .cells = malloc(desc->cell_count.x*desc->cell_count.y*sizeof(Cell)),
     };
     return grid;
 }
@@ -41,8 +42,8 @@ void grid_insert(Grid *grid, Box box) {
     }
 }
 
-Vec(Box) grid_query(Grid grid, Box area) {
-    const Vec2 cell_size = vec2_div(grid.world_box.size, grid.cell_count);
+Vec(Box) grid_query(const Grid* grid, Box area) {
+    const Vec2 cell_size = vec2_div(grid->world_box.size, grid->cell_count);
 
     Vec2 top_left = vec2_div(area.pos, cell_size);
     top_left.x = floorf(top_left.x);
@@ -55,7 +56,7 @@ Vec(Box) grid_query(Grid grid, Box area) {
     Vec(Box) result = NULL;
     for (int32_t y = top_left.y; y < bottom_right.y; y++) {
         for (int32_t x = top_left.x; x < bottom_right.x; x++) {
-            Cell cell = grid.cells[x+y*(int) grid.cell_count.x];
+            Cell cell = grid->cells[x+y*(int) grid->cell_count.x];
             for (uint32_t i = 0; i < cell.box_i; i++) {
                 vec_push(result, cell.boxes[i]);
             }
@@ -65,18 +66,17 @@ Vec(Box) grid_query(Grid grid, Box area) {
     return result;
 }
 
-void grid_reset(Grid *grid) {
+void grid_clear(Grid *grid) {
     for (uint32_t i = 0; i < grid->cell_count.x*grid->cell_count.y; i++) {
         grid->cells[i].box_i = 0;
     }
 }
 
-void grid_debug_draw(Grid grid, SDL_Renderer *renderer) {
-    const Vec2 cell_size = vec2_div(grid.world_box.size, grid.cell_count);
[…]

### 2025-01-23 src/hashing.c

@@ -1,4 +1,5 @@
 #include "hashing.h"
+#include <SDL2/SDL_render.h>
 #include <stdlib.h>
 #include <math.h>
 #include <stdio.h>
@@ -12,13 +13,13 @@ static uint64_t hash_position(int32_t x, int32_t y) {
     return full;
 }
 
-SpatialHash spatial_hash_new(uint32_t map_capacity, Vec2 cell_size) {
-    SpatialHash space = {
-        .cell_size = cell_size,
-        .map_capacity = map_capacity,
-        .buckets = calloc(map_capacity, sizeof(Bucket)),
+SpatialHash* spatial_hash_new(const SpatialHashDesc* desc) {
+    SpatialHash* space = malloc(sizeof(SpatialHash));
+    *space = (SpatialHash) {
+        .cell_size = desc->cell_size,
+        .map_capacity = desc->map_capacity,
+        .buckets = calloc(desc->map_capacity, sizeof(Bucket)),
     };
-
     return space;
 }
 
@@ -54,21 +55,21 @@ void spatial_hash_clear(SpatialHash *space) {
     }
 }
 
-Vec(Box) spatial_hash_query(SpatialHash space, Box area) {
+Vec(Box) spatial_hash_query(const SpatialHash* space, Box area) {
     Vec(Box) result = NULL;
 
-    Vec2 min = vec2_div(area.pos, space.cell_size);
+    Vec2 min = vec2_div(area.pos, space->cell_size);
     min.x = floorf(min.x);
     min.y = floorf(min.y);
-    Vec2 max = vec2_div(vec2_add(area.pos, area.size), space.cell_size);
+    Vec2 max = vec2_div(vec2_add(area.pos, area.size), space->cell_size);
     max.x = ceilf(max.x);
     max.y = ceilf(max.y);
 
     for (int32_t y = min.y; y < max.y; y++) {
         for (int32_t x = min.x; x < max.x; x++) {
             uint64_t hash = hash_position(x, y);
-            uint64_t index = hash % space.map_capacity;
-            Bucket *bucket = &space.buckets[index];
+            uint64_t index = hash % space->map_capacity;
+            Bucket *bucket = &space->buckets[index];
             for (uint32_t i = 0; i < bucket->box_i; i++) {
                 vec_push(result, bucket->boxes[i]);
             }
@@ -78,20 +79,30 @@ Vec(Box) spatial_hash_query(SpatialHash space, Box area) {
     return result;
 }
 
[…]

### 2025-01-23 src/main.c

@@ -5,21 +5,21 @@
 #include "color.h"
 #include "hashing.h"
 #include "window.h"
-
-#include <SDL2/SDL_mouse.h>
-#include <SDL2/SDL_rect.h>
-#include <SDL2/SDL_render.h>
-#include <time.h>
-#include <unistd.h>
-
-#include <SDL2/SDL_timer.h>
-#include <stdint.h>
-#include <stdlib.h>
-
 #include "ds.h"
 #include "box.h"
 #include "quadtree.h"
 #include "grid.h"
+#include "strategy_interface.h"
+
+#include <time.h>
+#include <unistd.h>
+#include <stdint.h>
+#include <stdlib.h>
+
+#include <SDL2/SDL_mouse.h>
+#include <SDL2/SDL_rect.h>
+#include <SDL2/SDL_render.h>
+#include <SDL2/SDL_timer.h>
 
 static float frand(void) {
     return (float) rand() / RAND_MAX;
@@ -64,129 +64,48 @@ static void init_boxes(Vec(Box) *boxes, uint32_t count) {
     }
 }
 
-static void benchmark_naive(Window window) {
-    Benchmark *bm = NULL;
-
+static void benchmark(Window window, Strategy strat, const void* desc, const char* name) {
     for (uint32_t box_count = config.iter.init_box_count; box_count <= config.iter.max_box_count; box_count BOX_INCREASE) {
-        printf("Naive: Benchmarking %u boxes with %u iterations...\n", box_count, config.iter.count);
+        printf("%s: Benchmarking %u boxes with %u iterations...\n", name, box_count, config.iter.count);
 
         Vec(Box) boxes = NULL;
         init_boxes(&boxes, box_count);
 
-        Vec(double) iter_times = NULL;
-        for (size_t i = 0; i < config.iter.count; i++) {
-            window_clear(window, color_rgb_hex(0x000000));
+        void* data = strat.new(desc);
 
+        // Insert all the boxes into the space.
+        for (size_t i = 0; i < config.iter.count; i++) {
+            for (size_t i = 0; i < vec_len(boxes); i++) {
+                strat.insert(data, boxes[i]);
+            }
+
+            // Check for collisions.
             Vec(Box) colliding_boxes = NULL;
             Vec(Box) non_colliding_boxes = NULL;
-
-            bench_func(iter_time) {
-                for (uint32_t i = 0; i < box_count; i++) {
-                    bool collided = false;
-                    for (uint32_t j = 0; j < box_count; j++) {
[…]

### 2025-01-23 src/naive.c

@@ -1,0 +1,33 @@
+#include "naive.h"
+#include "ds.h"
+
+Naive* naive_new(const void* desc) {
+    (void) desc;
+    Naive* naive = malloc(sizeof(Naive));
+    *naive = (Naive) {0};
+    return naive;
+}
+
+void naive_free(Naive* data) {
+    vec_free(data->boxes);
+}
+
+void naive_insert(Naive* data, Box box) {
+    vec_push(data->boxes, box);
+}
+
+void naive_clear(Naive* data) {
+    vec_free(data->boxes);
+}
+
+Vec(Box) naive_query(const Naive* data, Box area) {
+    (void) area;
+    Vec(Box) result = NULL;
+    vec_insert_arr(result, 0, data->boxes, vec_len(data->boxes));
+    return result;
+}
+
+void naive_debug_draw(const Naive* data, SDL_Renderer* renderer) {
+    (void) data;
+    (void) renderer;
+}

### 2025-01-23 src/quadtree.c

@@ -83,17 +83,18 @@ static void quadtree_node_insert(Quadtree *quadtree, QuadtreeNode *node, Box box
     node->boxes[node->box_i++] = box;
 }
 
-Quadtree quadtree_new(Box area, int max_depth, int max_box_count) {
-    size_t max_node_count = powf(4, max_depth);
-    Quadtree quadtree = {
-        .max_depth = max_depth,
-        .max_box_count = max_box_count,
+Quadtree* quadtree_new(const QuadtreeDesc* desc) {
+    size_t max_node_count = powf(4, desc->max_depth);
+    Quadtree* quadtree = malloc(sizeof(Quadtree));
+    *quadtree = (Quadtree) {
+        .max_depth = desc->max_depth,
+        .max_box_count = desc->max_box_count,
         .node_pool = malloc(max_node_count * sizeof(QuadtreeNode)),
         .node_pool_i = 1,
     };
 
-    quadtree.node_pool[0] = (QuadtreeNode) {
-        .area = area,
+    quadtree->node_pool[0] = (QuadtreeNode) {
+        .area = desc->area,
     };
 
     return quadtree;
@@ -116,6 +117,7 @@ Quadtree quadtree_new(Box area, int max_depth, int max_box_count) {
 
 void quadtree_free(Quadtree *quadtree) {
     free(quadtree->node_pool);
+    free(quadtree);
 }
 
 void quadtree_insert(Quadtree *quadtree, Box box) {
@@ -129,7 +131,7 @@ void quadtree_clear(Quadtree *quadtree) {
     };
 }
 
-void quadtree_query_helper(QuadtreeNode *node, Box area, Vec(Box) *result) {
+void quadtree_query_helper(const QuadtreeNode *node, Box area, Vec(Box) *result) {
     if (node == NULL || !box_overlapp(node->area, area)) {
         return;
     }
@@ -142,13 +144,13 @@ void quadtree_query_helper(QuadtreeNode *node, Box area, Vec(Box) *result) {
     vec_insert_arr(*result, vec_len(*result), node->boxes, node->box_i);
 }
 
-Vec(Box) quadtree_query(Quadtree quadtree, Box area) {
+Vec(Box) quadtree_query(const Quadtree* quadtree, Box area) {
     Vec(Box) result = NULL;
-    quadtree_query_helper(&quadtree.node_pool[0], area, &result);
[…]

### 2025-01-23 include/benchmark.h
Beginning of benchmark refactor

@@ -9,24 +9,20 @@ typedef struct Benchmark Benchmark;
 struct Benchmark {
     // Stack of benchmarks.
     Benchmark *next;
-    uint32_t box_count;
-    Vec(double) iter_times;
 
+    const char* name;
+    uint32_t run_count;
     double total_ms;
     double average_ms;
     double min_ms;
     double max_ms;
+    HashMap(const char*, Benchmark*) children;
 };
 
 extern double get_time(void);
 
-#define bench_func(t) \
-    double t = 0.0; \
-    for (bool _i_ = (t = get_time(), false); !_i_; _i_ = (t = get_time() - t, true))
+extern void bm_begin(const char* fmt, ...) __attribute__((format(printf, 1, 2)));
+extern void bm_end(void);
+#define bm(fmt, ...) for (bool _i_ = (bm_begin(name, ##__VA_ARGS__), false); !_i_; _i_ = (bm_end(), true))
 
-// Pushes a benchmark onto the stack.
-// Don't free 'iter_times' it's stored within the benchmark. It is freed when
-// 'benchmark_free' is called.
-extern void benchmark_register(Benchmark **benchmark, Vec(double) iter_times, uint32_t box_count);
-extern void benchmark_free(Benchmark *benchmark);
-extern void benchmark_write_json(Benchmark *benchmark, const char *filename);
+extern void bm_dump(void);

### 2025-01-23 include/ds.h

@@ -85,9 +85,9 @@ struct HashMapDesc {
     int (*cmp)(const void *, const void *, size_t);
 };
 #define hash_map_desc_default(map) (HashMapDesc) { \
-    .key_size = sizeof(map->key), \
-    .value_size = sizeof(map->value), \
-    .bucket_size = sizeof(*map), \
+    .key_size = sizeof(__typeof__(map->key)), \
+    .value_size = sizeof(__typeof__(map->value)), \
+    .bucket_size = sizeof(__typeof__(*map)), \
     .value_offset = offsetof(__typeof__(*map), value), \
     .hash = fvn1a_hash, \
     .cmp = memcmp, \

### 2025-01-23 src/benchmark.c

@@ -1,10 +1,11 @@
 #include "benchmark.h"
 #include "ds.h"
 
-#include <math.h>
+#include <string.h>
 #include <time.h>
 #include <stdlib.h>
 #include <stdio.h>
+#include <stdarg.h>
 
 // Get the unix time stamp in miliseconds.
 double get_time(void) {
@@ -19,76 +20,94 @@ double get_time(void) {
     return (double) (ts.tv_sec-init_ts.tv_sec) * 1e3 + (double) (ts.tv_nsec-init_ts.tv_nsec) / 1e6;
 }
 
-void benchmark_free(Benchmark *benchmark) {
-    if (benchmark == NULL) {
+static size_t str_hash(const void* data, size_t len) {
+    (void) len;
+    char* const* str = data;
+    return fvn1a_hash(*str, strlen(*str));
+}
+
+static int str_cmp(const void* a, const void* b, size_t len) {
+    (void) len;
+    char* const* _a = a;
+    char* const* _b = b;
+    return strcmp(*_a, *_b);
+}
+
+Benchmark root = {0};
+Benchmark *curr = NULL;
+
+void bm_begin(const char* fmt, ...) {
+    HashMapDesc desc = hash_map_desc_default(root.children);
+    desc.hash = str_hash;
+    desc.cmp = str_cmp;
+
+    va_list args;
+    va_start(args, fmt);
+    int len = vsnprintf(NULL, 0, fmt, args);
+    char* name = malloc(len + 1);
+    vsnprintf(name, len + 1, fmt, args);
+    va_end(args);
+    // printf("Name: %s\n", name);
+
+    if (root.children == NULL) {
+        hash_map_new(root.children, desc);
+    }
+
+    Benchmark* bm;
+    if (curr == NULL) {
+        bm = hash_map_get(root.children, name);
+    } else {
+        bm = hash_map_get(curr->children, name);
+    }
+
+    // New benchmark
+    if (bm == NULL) {
+        bm = malloc(sizeof(Benchmark));
+        *bm = (Benchmark) {
+            .name = name,
+        };
+        hash_map_new(bm->children, desc);
+        if (curr == NULL) {
+            hash_map_insert(root.children, name, bm);
+        } else {
+            hash_map_insert(curr->children, name, bm);
+        }
+        printf("New bm: %s\n", name);
+    } else {
+        free(name);
+    }
+
+    bm->run_count++;
+
+    bm->next = curr;
+    curr = bm;
+}
+
[…]

### 2025-01-23 src/grid.c

@@ -10,7 +10,7 @@ Grid* grid_new(const GridDesc* desc) {
     *grid = (Grid) {
         .world_box = desc->grid_size,
         .cell_count = desc->cell_count,
-        .cells = malloc(desc->cell_count.x*desc->cell_count.y*sizeof(Cell)),
+        .cells = calloc(desc->cell_count.x*desc->cell_count.y, sizeof(Cell)),
     };
     return grid;
 }

### 2025-01-23 src/main.c

@@ -11,6 +11,7 @@
 #include "grid.h"
 #include "strategy_interface.h"
 
+#include <stdio.h>
 #include <time.h>
 #include <unistd.h>
 #include <stdint.h>
@@ -42,7 +43,7 @@ struct Config {
 const Config config = {
     .box_size = 4,
     .iter = {
-        .count = 32,
+        .count = 1,
         .init_box_count = 0,
         .max_box_count = 1000,
     },
@@ -64,8 +65,10 @@ static void init_boxes(Vec(Box) *boxes, uint32_t count) {
     }
 }
 
-static void benchmark(Window window, Strategy strat, const void* desc, const char* name) {
+static void run(Window window, Strategy strat, const void* desc, const char* name) {
     for (uint32_t box_count = config.iter.init_box_count; box_count <= config.iter.max_box_count; box_count BOX_INCREASE) {
+        bm_begin("box-%u", box_count);
+
         printf("%s: Benchmarking %u boxes with %u iterations...\n", name, box_count, config.iter.count);
 
         Vec(Box) boxes = NULL;
@@ -73,18 +76,25 @@ static void benchmark(Window window, Strategy strat, const void* desc, const cha
 
         void* data = strat.new(desc);
 
-        // Insert all the boxes into the space.
         for (size_t i = 0; i < config.iter.count; i++) {
+            // Insert all the boxes into the space.
+            bm_begin("insert");
             for (size_t i = 0; i < vec_len(boxes); i++) {
                 strat.insert(data, boxes[i]);
             }
+            bm_end();
 
             // Check for collisions.
+            bm_begin("collision");
             Vec(Box) colliding_boxes = NULL;
             Vec(Box) non_colliding_boxes = NULL;
             for (size_t i = 0; i < vec_len(boxes); i++) {
-                bool collided = false;
+                // Query
+                bm_begin("query");
                 Vec(Box) near = strat.query(data, boxes[i]);
+                bm_end();
+
+                bool collided = false;
                 for (size_t j = 0; j < vec_len(near); j++) {
                     if (box_eq(boxes[i], near[j])) {
                         continue;
@@ -101,6 +111,7 @@ static void benchmark(Window window, Strategy strat, const void* desc, const cha
                     vec_push(non_colliding_boxes, boxes[i]);
[…]

### 2025-01-23 include/benchmark.h
Finished benchmark refactor.

@@ -17,6 +17,8 @@ struct Benchmark {
     double min_ms;
     double max_ms;
     HashMap(const char*, Benchmark*) children;
+
+    double _start;
 };
 
 extern double get_time(void);
@@ -26,3 +28,4 @@ extern void bm_end(void);
 #define bm(fmt, ...) for (bool _i_ = (bm_begin(name, ##__VA_ARGS__), false); !_i_; _i_ = (bm_end(), true))
 
 extern void bm_dump(void);
+extern void bm_dump_json(const char* filename);

### 2025-01-23 src/benchmark.c

@@ -6,6 +6,7 @@
 #include <stdlib.h>
 #include <stdio.h>
 #include <stdarg.h>
+#include <float.h>
 
 // Get the unix time stamp in miliseconds.
 double get_time(void) {
@@ -43,11 +44,13 @@ void bm_begin(const char* fmt, ...) {
 
     va_list args;
     va_start(args, fmt);
-    int len = vsnprintf(NULL, 0, fmt, args);
+    va_list args_len;
+    va_copy(args_len, args);
+    int len = vsnprintf(NULL, 0, fmt, args_len);
+    va_end(args_len);
     char* name = malloc(len + 1);
     vsnprintf(name, len + 1, fmt, args);
     va_end(args);
-    // printf("Name: %s\n", name);
 
     if (root.children == NULL) {
         hash_map_new(root.children, desc);
@@ -65,6 +68,7 @@ void bm_begin(const char* fmt, ...) {
         bm = malloc(sizeof(Benchmark));
         *bm = (Benchmark) {
             .name = name,
+            .min_ms = DBL_MAX,
         };
         hash_map_new(bm->children, desc);
         if (curr == NULL) {
@@ -72,18 +76,28 @@ void bm_begin(const char* fmt, ...) {
         } else {
             hash_map_insert(curr->children, name, bm);
         }
-        printf("New bm: %s\n", name);
     } else {
         free(name);
     }
 
     bm->run_count++;
+    bm->_start = get_time();
 
     bm->next = curr;
     curr = bm;
 }
 
 void bm_end(void) {
+    double time = get_time() - curr->_start;
+    curr->total_ms += time;
+    curr->average_ms = curr->total_ms / curr->run_count;
+    if (curr->min_ms > time) {
+        curr->min_ms = time;
+    }
+    if (curr->max_ms < time) {
+        curr->max_ms = time;
+    }
+
     curr = curr->next;
 }
 
@@ -94,7 +108,11 @@ void bm_dump_helper(const Benchmark* bm, int depth) {
 
     char spaces[512] = {0};
     memset(spaces, ' ', 4 * depth);
-    printf("%s%s\n", spaces, bm->name);
+    printf("%s%s: (runs: %d), (total: %.2f ms)\n",
+        spaces,
+        bm->name,
+        bm->run_count,
+        bm->average_ms);
 
     HashMapIter iter = hash_map_iter_new(bm->children);
[…]

### 2025-01-23 src/main.c

@@ -43,7 +43,7 @@ struct Config {
 const Config config = {
     .box_size = 4,
     .iter = {
-        .count = 1,
+        .count = 32,
         .init_box_count = 0,
         .max_box_count = 1000,
     },
@@ -67,7 +67,7 @@ static void init_boxes(Vec(Box) *boxes, uint32_t count) {
 
 static void run(Window window, Strategy strat, const void* desc, const char* name) {
     for (uint32_t box_count = config.iter.init_box_count; box_count <= config.iter.max_box_count; box_count BOX_INCREASE) {
-        bm_begin("box-%u", box_count);
+        bm_begin("%u", box_count);
 
         printf("%s: Benchmarking %u boxes with %u iterations...\n", name, box_count, config.iter.count);
 
@@ -197,7 +197,8 @@ int32_t main(void) {
     run(window, STRATEGY_SPATIAL_HASHING, &sh_desc, "Spatial Hashing");
     bm_end();
 
-    bm_dump();
+    // bm_dump();
+    bm_dump_json("benchmark.json");
 
     window_destroy(&window);
     return 0;

### 2025-01-23 include/benchmark.h
Benchmark using different distribution functions.

@@ -29,3 +29,4 @@ extern void bm_end(void);
 
 extern void bm_dump(void);
 extern void bm_dump_json(const char* filename);
+extern void bm_reset(void);

### 2025-01-23 src/benchmark.c

@@ -196,3 +196,8 @@ void bm_dump_json(const char* filename) {
     fprintf(fp, "}");
     fclose(fp);
 }
+
+void bm_reset(void) {
+    hash_map_free(root.children);
+    curr = NULL;
+}

### 2025-01-23 src/main.c

@@ -54,28 +54,58 @@ const Config config = {
 };
 #define BOX_INCREASE += 10
 
-static void init_boxes(Vec(Box) *boxes, uint32_t count) {
+typedef void (*RandomPointsFunc)(Vec2* points, int count, Vec2 space);
+
+void even_distribution(Vec2* points, int count, Vec2 space) {
+    for (int i = 0; i < count; i++) {
+        points[i] = vec2_mul(vec2(frand(), frand()), space);
+    }
+}
+
+// Probability density function
+float pdf(Vec2 point) {
+    Vec2 c = vec2s(0.5f);
+    Vec2 delta = vec2_sub(point, c);
+    // Gaussian falloff
+    return expf(-(delta.x * delta.x + delta.y * delta.y) / 0.1f);
+}
+
+void uneven_distribution(Vec2* points, int count, Vec2 space) {
+    int i = 0;
+    while (i < count) {
+        Vec2 point = vec2(frand(), frand());
+        float p = pdf(point);
+        if (frand() < p) {
+            points[i++] = vec2_mul(point, space);
+        }
+    }
+}
+
+static void init_boxes(Vec(Box) *boxes, uint32_t count, RandomPointsFunc rand_points_func) {
     srand(1234);
+
+    Vec2* pos = malloc(sizeof(Vec2) * count);
+    rand_points_func(pos, count, vec2(config.world.width - config.box_size, config.world.height - config.box_size));
     for (size_t i = 0; i < count; i++) {
         Box box = {
-            .pos = vec2(frand()*(config.world.width-config.box_size), frand()*(config.world.height-config.box_size)),
+            .pos = pos[i],
             .size = vec2s(config.box_size),
         };
         vec_push(*boxes, box);
     }
+    free(pos);
 }
 
-static void run(Window window, Strategy strat, const void* desc, const char* name) {
+static void run(Window window, Strategy strat, const void* desc, const char* name, RandomPointsFunc rand_points_func) {
     for (uint32_t box_count = config.iter.init_box_count; box_count <= config.iter.max_box_count; box_count BOX_INCREASE) {
-        bm_begin("%u", box_count);
-
         printf("%s: Benchmarking %u boxes with %u iterations...\n", name, box_count, config.iter.count);
 
         Vec(Box) boxes = NULL;
-        init_boxes(&boxes, box_count);
[…]

### 2025-01-23 graph.py
Improved graph generating script used to generate all the graphs from the data.

@@ -1,39 +1,38 @@
 import matplotlib.pyplot as plt
 import json
+import collections
+import sys
 
-class Data:
-    def __init__(self, name, file) -> None:
-        self.file = file
-        self.name = name
-        with open(file) as fp:
-            self.data = json.load(fp)
+if len(sys.argv) != 3:
+    print("Please pass in a benchmark JSON file and a file prefix.")
+    print(f"{sys.argv[0]} [benchmark] [prefix]")
+    exit(1)
 
-class Graph:
-    def __init__(self) -> None:
-        self.fig, self.ax = plt.subplots()
-        self.ax.set(xlabel="box count", ylabel="time (ms)")
-        self.ax.grid()
+with open(sys.argv[1], "rb") as f:
+    data = json.load(f)
 
-    def plot_data(self, data: Data) -> None:
-        x = []
-        y = []
-        for benchmark in data.data:
-            x.append(benchmark["box_count"])
-            y.append(benchmark["average"])
-        self.ax.plot(x, y, label=data.name);
+def graph(name, data_point):
+    fig, ax = plt.subplots()
+    ax.set(xlabel="box count", ylabel="time (ms)")
+    ax.grid()
+    for strat in data:
+        print(f"{strat}: {sys.argv[2]}-{name}.png")
+        run_times = {}
+        for box_count in data[strat]["children"]:
+            run_times[int(box_count)] = float(data_point(data[strat]["children"][box_count]))
+
+        sorted_run_times = collections.OrderedDict(sorted(run_times.items()))
+        ax.plot(list(sorted_run_times.keys()), list(sorted_run_times.values()), label=strat)
         plt.legend()
 
-    def save(self, filepath: str):
-        self.fig.savefig(filepath)
+    fig.savefig(f"graphs/{sys.argv[2]}-{name}.png")
 
-data = [
-    Data("naive", "naive.json"),
-    Data("grid", "grid.json"),
-    Data("quadtree", "quadtree.json"),
-    Data("spatial hashing", "spatial_hashing.json"),
-]
-
-graph = Graph()
-for d in data:
-    graph.plot_data(d)
-graph.save("graph.png")
+graph("total", lambda box: box["total"])
+graph("insert", lambda box: box["children"]["insert"]["total"])
+graph("collision", lambda box: box["children"]["collision"]["total"])
+def query(box):
+    children = box["children"]["collision"]["children"]
+    if "query" in children:
+        return children["query"]["total"]
+    return 0.0
+graph("query", query)

### 2025-01-23 paper/paper.tex
Abstract has been writen and inserted image for quadtree.

@@
 
-\usepackage[utf8]{inputenc} % For UTF-8 encoding
-\usepackage[english]{babel} % For English language settings
-% \usepackage{amsmath}        % For mathematical symbols
-% \usepackage{graphicx}       % For including images
-% \usepackage{hyperref}       % For hyperlinks
-\usepackage[margin=1in]{geometry}       % To set margins
-% \usepackage{float}          % For controlling figure placement
-% \usepackage{booktabs}       % For better table formatting
+\usepackage[utf8]{inputenc}
+\usepackage[english]{babel}
+\usepackage{graphicx}
+\usepackage[margin=1in]{geometry}
+\usepackage{svg}
 
 \renewcommand{\baselinestretch}{1.15}
-
 \setlength{\parindent}{0pt}
@@
 
-\title{Comparison of Spatial Partitioning Strategies}
-\author{Linus Kåreblom}
-\date{\today}
+\graphicspath{ {../graphs/} }
 
@@
 \begin{abstract}
-    This is the abstract of my paper.
+    Spatial partitioning is a widely used optimization strategy to efficiently
+    optimize objects in space. It can be found in many applications, from
+    simulating galaxies to video games.
+    Checking collision between objects is an expensive operation that quickly
+    becomes a bottleneck as the number of objects grow.
+    This paper compares three different strategies, alongside a naive solution.
+    The aim is to evaluate the use cases for the different strategies.
+    Each strategy was implemented to fit the same interface to make it as fair
+    of a comparison as possible.
+    The results show how spatial hashing out performs every other method when
+    the objects are distributed evenly while quadtrees excel with uneven object
+    distribution.
 \end{abstract}
@@
 The data structure is comprised of nodes. Each node has four pointers to other
-nodes, could also be referred to as branches.
+nodes, could also be referred to as branches or leaves. This strategy works by
+inserting points into a leaf until the leaf reaches its capacity, then it splits
+into four equally large sub leaves. This subdivision pattern is shown in figure
+\ref{fig:quadtree}.
+
+\begin{figure}[ht]
+    \includesvg[width=\textwidth]{images/invert-quadtree.svg}
+    \caption{Visual representation of a quadtree.}
+    \label{fig:quadtree}
+\end{figure}
 

### 2025-01-23 graph.py
Generated graphs with more data points and excluded the naive solution.

@@ -11,11 +11,14 @@ if len(sys.argv) != 3:
 with open(sys.argv[1], "rb") as f:
     data = json.load(f)
 
-def graph(name, data_point):
+def graph(name, data_point, include_naive=False):
     fig, ax = plt.subplots()
     ax.set(xlabel="box count", ylabel="time (ms)")
     ax.grid()
     for strat in data:
+        if not include_naive and strat == "naive":
+            continue;
+
         print(f"{strat}: {sys.argv[2]}-{name}.png")
         run_times = {}
         for box_count in data[strat]["children"]:
@@ -27,12 +30,14 @@ def graph(name, data_point):
 
     fig.savefig(f"graphs/{sys.argv[2]}-{name}.png")
 
-graph("total", lambda box: box["total"])
-graph("insert", lambda box: box["children"]["insert"]["total"])
-graph("collision", lambda box: box["children"]["collision"]["total"])
+graph("average", lambda box: box["average"])
+graph("insert", lambda box: box["children"]["insert"]["average"])
+graph("collision", lambda box: box["children"]["collision"]["average"])
 def query(box):
     children = box["children"]["collision"]["children"]
     if "query" in children:
-        return children["query"]["total"]
+        return children["query"]["average"]
     return 0.0
 graph("query", query)
+
+# graph("naive", lambda box: box["average"], True)

### 2025-01-23 src/benchmark.c

@@ -146,16 +146,16 @@ void bm_dump_json_helper(FILE* fp, const Benchmark* bm, int depth) {
     fprintf(fp, "%s\"run_count\": %u,\n", spaces, bm->run_count);
 
     // Average time
-    fprintf(fp, "%s\"average\": %f,\n", spaces, bm->average_ms);
+    fprintf(fp, "%s\"average\": %.15f,\n", spaces, bm->average_ms);
 
     // Total time
-    fprintf(fp, "%s\"total\": %f,\n", spaces, bm->total_ms);
+    fprintf(fp, "%s\"total\": %.15f,\n", spaces, bm->total_ms);
 
     // Min time
-    fprintf(fp, "%s\"min\": %f,\n", spaces, bm->min_ms);
+    fprintf(fp, "%s\"min\": %.15f,\n", spaces, bm->min_ms);
 
     // Max time
-    fprintf(fp, "%s\"max\": %f,\n", spaces, bm->max_ms);
+    fprintf(fp, "%s\"max\": %.15f,\n", spaces, bm->max_ms);
 
     // Children
     fprintf(fp, "%s\"children\": {\n", spaces);

### 2025-01-23 src/main.c

@@ -41,11 +41,11 @@ struct Config {
 };
 
 const Config config = {
-    .box_size = 4,
+    .box_size = 2,
     .iter = {
         .count = 32,
-        .init_box_count = 0,
-        .max_box_count = 1000,
+        .init_box_count = 10,
+        .max_box_count = 2500,
     },
     .world = {
         .width = 1280,
@@ -96,7 +96,7 @@ static void init_boxes(Vec(Box) *boxes, uint32_t count, RandomPointsFunc rand_po
     free(pos);
 }
 
-static void run(Window window, Strategy strat, const void* desc, const char* name, RandomPointsFunc rand_points_func) {
+static void run(Window* window, Strategy strat, const void* desc, const char* name, RandomPointsFunc rand_points_func) {
     for (uint32_t box_count = config.iter.init_box_count; box_count <= config.iter.max_box_count; box_count BOX_INCREASE) {
         printf("%s: Benchmarking %u boxes with %u iterations...\n", name, box_count, config.iter.count);
 
@@ -144,11 +144,11 @@ static void run(Window window, Strategy strat, const void* desc, const char* nam
             bm_end();
 
             // Visualize.
-            window_clear(window, color_rgb_hex(0x000000));
-            SDL_SetRenderDrawColor(window.renderer, 64, 64, 64, 255);
-            strat.debug_draw(data, window.renderer);
+            window_clear(*window, color_rgb_hex(0x000000));
+            SDL_SetRenderDrawColor(window->renderer, 64, 64, 64, 255);
+            strat.debug_draw(data, window->renderer);
 
-            SDL_SetRenderDrawColor(window.renderer, 255, 0, 0, 255);
+            SDL_SetRenderDrawColor(window->renderer, 255, 0, 0, 255);
             for (size_t i = 0; i < vec_len(colliding_boxes); i++) {
                 SDL_FRect rect = {
                     .x = colliding_boxes[i].pos.x,
@@ -156,10 +156,10 @@ static void run(Window window, Strategy strat, const void* desc, const char* nam
                     .w = colliding_boxes[i].size.w,
                     .h = colliding_boxes[i].size.h,
                 };
-                SDL_RenderDrawRectF(window.renderer, &rect);
+                SDL_RenderDrawRectF(window->renderer, &rect);
             }
 
-            SDL_SetRenderDrawColor(window.renderer, 255, 255, 255, 255);
[…]

### 2025-01-23 paper/paper.tex
Paper: Material and Method and start writing Result.

@@
 
+\section{Material and Method}
+
+\subsection{Linux}
+Linux is the operating system used to develop the application and write this
+paper. The program uses a specific Linux API to acquire the current time as a
+UNIX time stamp. The way to do this would differ on other operating systems.
+
+\subsection{Neovim}
+Neovim was the editor used to edit the code. It is a modal text editor and not a
+full integrated development environment, or IDE as they are called. A plugin
+which utilizes the language server protocol was used to provide me with
+intellisense and to aid in development by warning me about syntax errors.
+
+\subsection{Git}
+Git is a version control system which helps track changes to the project. In
+this project it is utilized to keep a time line of the development process and
+make sure changes could be synced between devices. The local git repository is
+mirrored on GitHub making it accessible from other devices, and the source code
+can be read from the browser.
+
+\subsection{CMake}
+CMake is a buildsystem, used to generate generate the final executable from the
+source code. It aids in compiling and linking the program making cross
+compilation and dependency management trivial.
+
+\subsection{C}
+C was the program of choice for this project due to its low level control.
+Since memory is managed manually in C I could be sure the results would not be
+skewed by a garbage collector running in the middle of a benchmark.
+
+\subsection{SDL2}
+SDL2 - Simple DirectMedia Layer - is a cross-platform library supplying an easy
+API to create a window and render to it. This was crucial when developing since
+visualizing the data structures made it easier to spot errors than simply
+looking at the code. It also gives a nice visual representation of the program.
+
+\subsection{Python}
+Python is an interpreted programming language which was used for making a script
+to generate graphs. All the graphs used in this paper was generated from data
+collected from a C program, which dumped said data into a JSON file, which was
+then read by a Python script using the matplotlib library to generate nice
+looking graphs.
+
+\newpage
+
+\section{Results}
+This section presents the results of the benchmarking done on the three
+strategies - quadtrees, spatial hashing and static grids. The metrics measured
+was average runtime: the sum of all operations needed to run the program.
+Insertions: how long it took to insert all the objects into the data structure.
+Collision testing: total time spent checking objects against each other, which
+includes querying. Querying: average time spent retrieving objects from the data
+structure. Clearing: how long it takes to reset the data structure making it
+ready for rebuilding. Each metric was measured with a range of object count,
+labeled as 'box count' in the graphs, from ten to 2500 with an increase of ten
+objects between each measurement. At each stage was run 32 times and the results
+were averaged to give smoother graphs. Two benchmarks was run, one with an even
+distribution of objects and one with an uneven distribution.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.75\textwidth]{even-naive.png}
+    \caption{Runtime of naive solution and partitioning strategies.}
+    \label{fig:naive}
+\end{figure}
+
+Figure \ref{fig:naive} shows a drastic difference in runtime between the naive
+solution and the optimization strategies. It is hard to evaluate which strategy
+performs best due to the stark difference between the naive and the optimized
+versions. Due to this discrepancy the naive solution will not be included in the
+other graphs presented. This comparison was only run with 1000 objects due to
+the time taken per execution.
+
+% \begin{figure}[h]
+%     \centering
+%     \includegraphics[width=0.75\textwidth]{even-average.png}
+%     \caption{Average runtime of spatial partitioning strategies.}
+%     \label{fig:even-average}
+% \end{figure}
+
 \end{document}

### 2025-01-23 src/main.c
Disable rendering codepath which could be skewing results.

@@ -22,6 +22,8 @@
 #include <SDL2/SDL_render.h>
 #include <SDL2/SDL_timer.h>
 
+static bool run_visually = true;
+
 static float frand(void) {
     return (float) rand() / RAND_MAX;
 }
@@ -45,7 +47,7 @@ const Config config = {
     .iter = {
         .count = 32,
         .init_box_count = 10,
-        .max_box_count = 2500,
+        .max_box_count = 1000,
     },
     .world = {
         .width = 1280,
@@ -144,36 +146,39 @@ static void run(Window* window, Strategy strat, const void* desc, const char* na
             bm_end();
 
             // Visualize.
-            window_clear(*window, color_rgb_hex(0x000000));
-            SDL_SetRenderDrawColor(window->renderer, 64, 64, 64, 255);
-            strat.debug_draw(data, window->renderer);
+            if (run_visually) {
+                window_clear(*window, color_rgb_hex(0x000000));
+                SDL_SetRenderDrawColor(window->renderer, 64, 64, 64, 255);
+                strat.debug_draw(data, window->renderer);
 
-            SDL_SetRenderDrawColor(window->renderer, 255, 0, 0, 255);
-            for (size_t i = 0; i < vec_len(colliding_boxes); i++) {
-                SDL_FRect rect = {
-                    .x = colliding_boxes[i].pos.x,
-                    .y = colliding_boxes[i].pos.y,
-                    .w = colliding_boxes[i].size.w,
-                    .h = colliding_boxes[i].size.h,
-                };
-                SDL_RenderDrawRectF(window->renderer, &rect);
-            }
+                SDL_SetRenderDrawColor(window->renderer, 255, 0, 0, 255);
+                for (size_t i = 0; i < vec_len(colliding_boxes); i++) {
+                    SDL_FRect rect = {
+                        .x = colliding_boxes[i].pos.x,
+                        .y = colliding_boxes[i].pos.y,
+                        .w = colliding_boxes[i].size.w,
+                        .h = colliding_boxes[i].size.h,
+                    };
+                    SDL_RenderDrawRectF(window->renderer, &rect);
+                }
 
-            SDL_SetRenderDrawColor(window->renderer, 255, 255, 255, 255);
-            for (size_t i = 0; i < vec_len(non_colliding_boxes); i++) {
-                SDL_FRect rect = {
-                    .x = non_colliding_boxes[i].pos.x,
[…]

### 2025-01-24 paper/paper.tex
Paper: Results section.

@@
     \centering
-    \includegraphics[width=0.75\textwidth]{even-naive.png}
+    \includegraphics[width=0.5\textwidth]{even-naive.png}
     \caption{Runtime of naive solution and partitioning strategies.}
@@
 
-% \begin{figure}[h]
-%     \centering
-%     \includegraphics[width=0.75\textwidth]{even-average.png}
-%     \caption{Average runtime of spatial partitioning strategies.}
-%     \label{fig:even-average}
-% \end{figure}
+\subsection{Average Runtime}
+
+\subsubsection{Evenly Distributed Objects}
+Figure \ref{fig:even-average} shows the average runtime for the program
+using the different strategies with an even object distribution. This includes
+all other measuring metrics. The X axis describes the amount of objects being
+processed. The Y axis describes how much time was spent in milliseconds
+executing the particular step in the program.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.5\textwidth]{even-average.png}
+    \caption{Average runtime of spatial partitioning strategies with evenly
+    distributed objects.}
+    \label{fig:even-average}
+\end{figure}
+
+\subsubsection{Unevenly Distributed Objects}
+Figure \ref{fig:uneven-average} presents the average runtime for the program
+using the different strategies with an uneven object distribution. This includes
+all other measuring metrics. As the object count increases the spatial hashing
+trends towards a $O(n^2)$ runtime complexity.
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.5\textwidth]{uneven-average.png}
+    \caption{Average runtime of spatial partitioning strategies with unevenly
+    distributed objects.}
+    \label{fig:uneven-average}
+\end{figure}
+
+\subsection{Insertion}
+
+\subsubsection{Evenly Distributed Objects}
+The runtime complexity of the insertion algorithms for evenly distributed
+objects is shown in figure \ref{fig:even-insert}.
+\begin{figure}[h]
+    \centering
+    \caption{Average insertion time per object with evenly distributed objects.}
+    \includegraphics[width=0.5\textwidth]{even-insert.png}
+    \label{fig:even-insert}
+\end{figure}
+
+\subsubsection{Unevenly Distributed Objects}
+The runtime complexity of the insertion algorithms for evenly distributed
+objects is shown in figure \ref{fig:uneven-insert}.
+\begin{figure}[h]
+    \centering
+    \caption{Average insertion time per object with unevenly distributed
+    objects.}
+    \includegraphics[width=0.5\textwidth]{uneven-insert.png}
+    \label{fig:uneven-insert}
+\end{figure}
+
+\subsection{Collision}
+
+\subsubsection{Evenly Distributed Objects}
+Figure \ref{fig:even-collision} shows the total time spent checking for
+colliding objects per iteration with an even distribution. This time also
+includes the time taken for querying.
+\begin{figure}[h]
+    \centering
+    \caption{Average collision time for all objects with evenly distributed
+    objects.}
+    \includegraphics[width=0.5\textwidth]{even-collision.png}
+    \label{fig:even-collision}
+\end{figure}
+
+\subsubsection{Unevenly Distributed Objects}
+Figure \ref{fig:uneven-collision} shows the total time spent checking for
+colliding objects per iteration with an uneven distribution. This time also
+includes the time taken for querying.
+\begin{figure}[h]
+    \centering
+    \caption{Average collision time for all objects with unevenly distributed
+    objects.}
+    \includegraphics[width=0.5\textwidth]{uneven-collision.png}
+    \label{fig:uneven-collision}
+\end{figure}
+
+\subsection{Querying}
+
+\subsubsection{Evenly Distributed Objects}
+Figure \ref{fig:even-query} represents the average time taken to query the space
+per object with an even distribution.
+\begin{figure}[h]
+    \centering
+    \caption{Average query time per object with evenly distributed objects.}
+    \includegraphics[width=0.5\textwidth]{even-query.png}
+    \label{fig:even-query}
+\end{figure}
+
+\subsubsection{Unevenly Distributed Objects}
+Figure \ref{fig:uneven-query} represents the average time taken to query the space
+per object with an uneven distribution.
+\begin{figure}[h]
+    \centering
+    \caption{Average query time per object with unevenly distributed objects.}
+    \includegraphics[width=0.5\textwidth]{uneven-query.png}
+    \label{fig:uneven-query}
+\end{figure}
 

### 2025-01-24 graph.py

@@ -13,7 +13,7 @@ with open(sys.argv[1], "rb") as f:
 
 def graph(name, data_point, include_naive=False):
     fig, ax = plt.subplots()
-    ax.set(xlabel="box count", ylabel="time (ms)")
+    ax.set(xlabel="Objects", ylabel="Time (ms)")
     ax.grid()
     for strat in data:
         if not include_naive and strat == "naive":

### 2025-01-24 src/main.c

@@ -47,7 +47,7 @@ const Config config = {
     .iter = {
         .count = 32,
         .init_box_count = 10,
-        .max_box_count = 1000,
+        .max_box_count = 2500,
     },
     .world = {
         .width = 1280,
@@ -220,7 +220,7 @@ int32_t main(void) {
             .grid_size = world_box,
             .cell_count = vec2(16, 16),
         };
-        bm_begin("grid");
+        bm_begin("Grid");
         run(window, STRATEGY_GRID, &grid_desc, "Grid", even_distribution);
         bm_end();
 
@@ -230,7 +230,7 @@ int32_t main(void) {
             .max_depth = 8,
             .max_box_count = 8,
         };
-        bm_begin("quadtree");
+        bm_begin("Quadtree");
         run(window, STRATEGY_QUADTREE, &qt_desc, "Quadtree", even_distribution);
         bm_end();
 
@@ -239,7 +239,7 @@ int32_t main(void) {
             .cell_size = vec2s(100.0f),
             .map_capacity = 4096,
         };
-        bm_begin("spatial-hashing");
+        bm_begin("Spatial Hashing");
         run(window, STRATEGY_SPATIAL_HASHING, &sh_desc, "Spatial Hashing", even_distribution);
         bm_end();
 
@@ -260,7 +260,7 @@ int32_t main(void) {
             .grid_size = world_box,
             .cell_count = vec2(16, 16),
         };
-        bm_begin("grid");
+        bm_begin("Grid");
         run(window, STRATEGY_GRID, &grid_desc, "Grid", uneven_distribution);
         bm_end();
 
@@ -270,7 +270,7 @@ int32_t main(void) {
             .max_depth = 8,
             .max_box_count = 8,
         };
-        bm_begin("quadtree");
+        bm_begin("Quadtree");
         run(window, STRATEGY_QUADTREE, &qt_desc, "Quadtree", uneven_distribution);
         bm_end();
 
@@ -279,7 +279,7 @@ int32_t main(void) {
             .cell_size = vec2s(100.0f),
             .map_capacity = 4096,
         };
-        bm_begin("spatial-hashing");
+        bm_begin("Spatial Hashing");
         run(window, STRATEGY_SPATIAL_HASHING, &sh_desc, "Spatial Hashing", uneven_distribution);
         bm_end();
 

### 2025-02-27 paper/paper.tex
paper: Wrote some analysis parts.

@@
 \usepackage{svg}
+\usepackage{float}
 
@@
 during the  collision detection phase. In this phase the program figures out
-which objects are close enough to be touching and overlapping. This wouldn't be
+which objects are close enough to be touching and overlapping. This would not be
 an issue in a small simulation with few objects, but I was striving for
@@
 an elegant solution to my problem. Diving deeper I discovered the three
-techniques I'll be comparing in this paper, along side the naive solution. The
+techniques I will be comparing in this paper, along side the naive solution. The
 basic theory of the technique is simply to group objects close to each other to
@@
 optimization to do query objects close to each other in space. During my
-research I found out about the three strategies I'm comparing in this paper. I'm
-doing this paper within the programming course.
+research I found out about the three strategies I am comparing in this paper. I
+am doing this paper within the programming course.
+
+\section{Purpose \& Scope}
+The purpose of this paper is to find the most optimal strategy for dividing up
+space to improve the speed of spatial querying of nearby objects.
+
+This study focuses mainly on CPU performance and runtime in exchange for less
+memory efficient implementation. Therefore time based evaluations of insertion,
+query time, clearing time and general collision testing time will be performed.
+
+The strategies are going to be evaluated on time based metrics - insertion,
+query, clearing and time spend doing collision testing. These evaluations are
+going to take place with two different distribution strategies of the objects
+though the same seed will be used for all tests so the pseudo-random numbers
+stay the same.
+
+These strategies could be extended into the third dimension but this study
+focuses solely focused on two dimensional space.
+
+The questions I'm going to answer are:
+\begin{itemize}
+    \item At what point is it worth implementing a more advanced data structure?
+    \item In what use case are the different algorithms most useful?
+\end{itemize}
 
@@
 Spatial partitioning is an optimization technique used to make programs run
-faster. It's often found within physics applications. The theory behind this
+faster. It is often found within physics applications. The theory behind this
 technique is to reduce the number of times one object is checked against
@@
 A quadtree is a recursive data structure, meaning it contains pointers to each
-objects of the same type as itself. It's in the family of tree data structures.
+objects of the same type as itself. It is in the family of tree data structures.
 The data structure is comprised of nodes. Each node has four pointers to other
@@
 slower because you have to request space within it from the operating system.
-Allocating is considered to be a slow operation since there's a lot of overhead
+Allocating is considered to be a slow operation since there is a lot of overhead
 introduced.
@@
 The static grid is built using a simple array. Indexing into the array is an
-$O(1)$ operation. The caveat with this approach is you're forced to know how
+$O(1)$ operation. The caveat with this approach is you are forced to know how
 big the area you want the to partition to be before compiling the program. Both
@@
 makes this approach super fast.
+
+\subsection{Hash Map}
+A hash map is a data structure which key value pairs allowing for aliasing of
+values. For example, a hash map using strings as keys and integers as values
+could map the key "life" to the value 42, meaning a query into the hash map for
+"life" would return 42. The way this works is by using a cryptographic hashing
+function to generate a unique hash for the object. This hash is a simple number
+which is usually very large. To make the hash useful you run modulo on it to
+make sure it's between zero and the number of buckets inside the hash map. The
+same method is used for retrieval, first running the key through the same
+hashing function, then wrapping it using modulo and returning what lies in the
+bucket. The only caveat with this is there could be collisions within the hash
+map. Two hashes could wrap into the same index. There are a multitude of ways to
+handle collisions like open addressing and separate chaining.
 
@@
 the object, then run a modulo operation on it giving me a number within the
-array bounds, and then it is inserted into the cell.
+array bounds, and then it is inserted into the cell. No collision resolution is
+taking place inside the structure meaning two cells could have the same bucket
+index which means queries for the one of the cells returns the objects in the
+other as well.
 
@@
 
-\section{Results}
-This section presents the results of the benchmarking done on the three
-strategies - quadtrees, spatial hashing and static grids. The metrics measured
-was average runtime: the sum of all operations needed to run the program.
-Insertions: how long it took to insert all the objects into the data structure.
-Collision testing: total time spent checking objects against each other, which
-includes querying. Querying: average time spent retrieving objects from the data
-structure. Clearing: how long it takes to reset the data structure making it
-ready for rebuilding. Each metric was measured with a range of object count,
-labeled as 'box count' in the graphs, from ten to 2500 with an increase of ten
-objects between each measurement. At each stage was run 32 times and the results
-were averaged to give smoother graphs. Two benchmarks was run, one with an even
-distribution of objects and one with an uneven distribution.
+\section{Results \& Analysis}
+This section presents the results and offers an analysis of the benchmarking
+done on the three strategies - quadtrees, spatial hashing and static grids. The
+metrics measured was average runtime: the sum of all operations needed to run
+the program. Insertions: how long it took to insert all the objects into the
+data structure. Collision testing: total time spent checking objects against
+each other, which includes querying. Querying: average time spent retrieving
+objects from the data structure. Clearing: how long it takes to reset the data
+structure making it ready for rebuilding. Each metric was measured with a range
+of object count, labeled as 'box count' in the graphs, from ten to 2500 with an
+increase of ten objects between each measurement. At each stage was run 32 times
+and the results were averaged to give smoother graphs. Two benchmarks was run,
+one with an even distribution of objects and one with an uneven distribution.
 
-\begin{figure}[h]
+\begin{figure}[H]
     \centering
@@
 
-\begin{figure}[h]
+\begin{figure}[H]
     \centering
@@
 trends towards a $O(n^2)$ runtime complexity.
-\begin{figure}[h]
+\begin{figure}[H]
     \centering
@@
 
+\subsubsection{Analysis}
+
+With evenly distributed objects The three strategies perform very similarly,
+though, spatial hashing does start showing more of an exponential runtime as the
+object count increases. Could be due to the implementation of the spatial
+hashing. Since it is using a hash map as its underlying data structure two
+different cells could have the same index within the same hash map meaning
+collision checks are happening for objects which are not close to each other in
+space. This happens because there is a limited amount of buckets available to
+store objects inside. The result of this scarcity is two hashes wrapping to the
+same bucket index. As we can see in \ref{fig:uneven-average} this issue becomes
+more apparent as the objects are less spread out.
+
 \subsection{Insertion}
@@
 The runtime complexity of the insertion algorithms for evenly distributed
-objects is shown in figure \ref{fig:even-insert}.
-\begin{figure}[h]
+objects is shown in figure \ref{fig:even-insert}. The metric being times is the
+insertion of all objects.
+\begin{figure}[H]
     \centering
@@
 The runtime complexity of the insertion algorithms for evenly distributed
-objects is shown in figure \ref{fig:uneven-insert}.
-\begin{figure}[h]
+objects is shown in figure \ref{fig:uneven-insert}. The metric being times is
+the insertion of all objects.
+\begin{figure}[H]
     \centering
@@
 
+\subsubsection{Analysis}
+There's virtually no difference in performance between the even and uneven
+insertion times. This is because the codepath taken for each insertion is the
+same, save for a different hash and bucket index. Because there is only a fixed
+amount of boxes able to be stored in each grid cell in both grid types the
+insertion time is run in $O(1)$, constant time. The case isn't the same for the
+quadtree since its a recursive algorithm. Each node has four sub-nodes which get
+traversed each insertion leading to a $O(\log_4 n)$ runtime since each step in
+the algorithm eliminates three of the sub-nodes. Once the correct node has been
+found the insertion time is a constant $O(1)$ since it uses the same technique
+as the other two.
+
+\subsection{Querying}
+
+\subsubsection{Evenly Distributed Objects}
+Figure \ref{fig:even-query} represents the average time taken to query the space
+per object with an even distribution.
+\begin{figure}[H]
+    \centering
+    \caption{Average query time per object with evenly distributed objects.}
+    \includegraphics[width=0.5\textwidth]{even-query.png}
+    \label{fig:even-query}
+\end{figure}
+
+\subsubsection{Unevenly Distributed Objects}
+Figure \ref{fig:uneven-query} represents the average time taken to query the space
+per object with an uneven distribution.
+\begin{figure}[H]
+    \centering
+    \caption{Average query time per object with unevenly distributed objects.}
+    \includegraphics[width=0.5\textwidth]{uneven-query.png}
+    \label{fig:uneven-query}
+\end{figure}
+
+\subsubsection{Analysis}
+
 \subsection{Collision}
@@
 includes the time taken for querying.
-\begin{figure}[h]
+\begin{figure}[H]
     \centering
@@
 
-\subsection{Querying}
+\section{Discussion}
 
-\subsubsection{Evenly Distributed Objects}
-Figure \ref{fig:even-query} represents the average time taken to query the space
-per object with an even distribution.
-\begin{figure}[h]
-    \centering
-    \caption{Average query time per object with evenly distributed objects.}
-    \includegraphics[width=0.5\textwidth]{even-query.png}
-    \label{fig:even-query}
-\end{figure}
-
-\subsubsection{Unevenly Distributed Objects}
-Figure \ref{fig:uneven-query} represents the average time taken to query the space
-per object with an uneven distribution.
-\begin{figure}[h]
-    \centering
-    \caption{Average query time per object with unevenly distributed objects.}
-    \includegraphics[width=0.5\textwidth]{uneven-query.png}
-    \label{fig:uneven-query}
-\end{figure}
+\section{Sources}
+Nystrom, Robert (2011) \textit{Game Programming Patterns}
 

### 2025-02-28 paper/paper.tex
paper: Change the title, abstraction and wrote the last analysis parts.

@@
 
-    {\Huge \textbf{Comparison of Spatial Partitioning Strategies}}\\[1.5cm]
+    {\Huge \textbf{Spatial Partitioning Strategies}}\\[1.5cm]
 
@@
     of a comparison as possible.
-    The results show how spatial hashing out performs every other method when
-    the objects are distributed evenly while quadtrees excel with uneven object
-    distribution.
+    The results show how quadtrees out performs every other method when
+    the objects are evenly and uneven distributed as the number of objects grow.
 \end{abstract}
@@
 
+\subsection{Vector}
+A vector refers to a dynamic array, not the mathematical concept of a vector.
+To allocate an array on the stack you need to know the size of it at compile
+time which means vectors have to live on the heap. The vector works by having a
+capacity which slowly fills up. Once it is full it resizes. This operation can
+be quite expensive since it needs to request new memory from the operating
+system and move all existing items into the new memory region.
+
 \section{Material and Method}
@@
 structure making it ready for rebuilding. Each metric was measured with a range
-of object count, labeled as 'box count' in the graphs, from ten to 2500 with an
+of object count, labeled as 'objects' in the graphs, from ten to 2500 with an
 increase of ten objects between each measurement. At each stage was run 32 times
@@
 same, save for a different hash and bucket index. Because there is only a fixed
-amount of boxes able to be stored in each grid cell in both grid types the
+amount of objects able to be stored in each grid cell in both grid types the
 insertion time is run in $O(1)$, constant time. The case isn't the same for the
@@
 \subsubsection{Analysis}
+In both cases the quadtree query has almost exactly the same time with the same
+$O(\log_4 n)$ runtime complexity. The other two, however, differ quite a bit
+time wise. This discrepancy is due to the use of vectors. Since a vector is used
+to return the overlapping objects the act of building this vector becomes more
+expensive as the amount of objects increases. Since there are more objects near
+each other in the unevenly distributed objects case there will naturally be more
+objects added to the vector. Quadtrees do not suffer from the same weakness
+since it has a limit on the amount of objects stored in each leaf meaning all
+queries has an upper limit to the number of objects, eliminating the need for
+vector resizing.
+
+The reason for the jaggedness in the graphs is because the execution time is so
+low the benchmark picks up on other things like the OS scheduler and other
+application doing work in the background taking away resources from the
+program.
 
@@
 
+\subsection{Analysis}
+Just like with querying the same thing happens with collisions. Because there
+are more objects close together there's a higher number of collision checks
+taking place. Since the quadtree limit the amount of objects in a cell the
+runtime doesn't change much between the two cases.
+
 \section{Discussion}
@@
 \section{Sources}
-Nystrom, Robert (2011) \textit{Game Programming Patterns}
+Nystom, Robert (2014) \textit{Game Programming Patterns}, genever benning
+
+Nystom, Robert (2021) \textit{Crafting Interpreters}, genever benning
 

### 2025-02-28 paper/paper.tex
paper: Finished the first draft.

@@
 \usepackage{float}
+\usepackage{hyperref}
 
@@
 
+\pagenumbering{alph}
 \begin{titlepage}
@@
 \end{titlepage}
+\pagenumbering{arabic}
 
@@
     optimize objects in space. It can be found in many applications, from
-    simulating galaxies to video games.
-    Checking collision between objects is an expensive operation that quickly
-    becomes a bottleneck as the number of objects grow.
-    This paper compares three different strategies, alongside a naive solution.
-    The aim is to evaluate the use cases for the different strategies.
-    Each strategy was implemented to fit the same interface to make it as fair
-    of a comparison as possible.
-    The results show how quadtrees out performs every other method when
-    the objects are evenly and uneven distributed as the number of objects grow.
+    simulating galaxies to video games. Checking collision between objects is an
+    expensive operation that quickly becomes a bottleneck as the number of
+    objects grow. This paper compares three different strategies, alongside a
+    naive solution. The aim is to evaluate the use cases for the different
+    strategies. Each strategy was implemented to fit the same interface to make
+    it as fair of a comparison as possible. The results show how quadtrees out
+    performs every other method when the objects are evenly and uneven
+    distributed as the number of objects grow.
 \end{abstract}
@@
 \section{Purpose \& Scope}
+\label{sec:purpose-and-scope}
 The purpose of this paper is to find the most optimal strategy for dividing up
@@
 memory efficient implementation. Therefore time based evaluations of insertion,
-query time, clearing time and general collision testing time will be performed.
+query time and general collision testing time will be performed.
 
 The strategies are going to be evaluated on time based metrics - insertion,
-query, clearing and time spend doing collision testing. These evaluations are
-going to take place with two different distribution strategies of the objects
-though the same seed will be used for all tests so the pseudo-random numbers
-stay the same.
+query and time spend doing collision testing. These evaluations are going to
+take place with two different distribution strategies of the objects though the
+same seed will be used for all tests so the pseudo-random numbers stay the same.
 
@@
 \section{Discussion}
+The design choices for this project was entirely focused on CPU performance
+leaving memory as an afterthought. This was done in order to mimic a modern game
+running on consumer hardware where memory is often an abundant resource.
+Everything inside the spatial partitioning strategies were allocated when the
+structure was initialized so no heap allocations were being issued in the middle
+of a hot loop. There are, however, two notable exception to this philosophy.
+When running queries I used vectors, which have to allocate memory on the heap,
+when doing queries. Looking back, there could be more efficient ways to handle
+this. Instead of having the query function handle the memory management, let the
+called handle that complexity. Have the caller allocate an array with sufficient
+enough size and pass that as a parameter for the function to modify. Though this
+would work it's not really a good solution, in my subjective opinion. Because
+there is guarantee of how many objects could be returned. The better solution
+would be to use a bump, arena or block allocator to further speed up this
+operation instead of using the general purpose malloc and free pattern provided
+by the C standard library. The other notable exception is the benchmarking tool
+I made for this. It uses a hash map to store how many times a function was run,
+the total time it ran and the average runtime per call. Because of my underlying
+implementation of the hash map it will require memory allocation when enough
+entries has been added. Not only that, but it was a nested structure which
+created a new hash map for each benchmark scope.
+
+The benchmarks are a thing I would have done differently if I would be redoing
+this project. There are a few decisions that I can't justify now. One of these
+weird choices is how I chose to benchmark insertions. Instead of running the
+benchmark for each object inserted I measured to total time taken to insert all
+objects. This doesn't sound weird until you look at how the rest of the
+benchmarks are measured. The queries are measured once per object instead of
+encapsulating all objects in a single benchmark. The insert metric is the only
+one which differs in this regard. It would be an easy fix code wise, but having
+to rewrite the paper based off of a new metric would be too much work to justify
+the change.
+
+I found that although quadtrees outperform all the other strategies in this
+particular program, it may not be the most suitable for all programs. Both grids
+and quadtrees limit you to a fixed known world size which could be a problem for
+larger games or simulations. If the quadtree is covering a massive region of the
+world then that would introduce significant overhead because there would have to
+be a lot of subdivisions happening. Chunking could fix this issue though there
+are other strategies more suitable for this. The spatial hashing grid solves
+this issue at the cost of performance. Grids come with the same world size
+restriction that the quadtree does but has the benefit of being extremely
+trivial to implement. At with almost all questions the answer to 'In what use
+case are the different algorithms most useful?' the answer is: it depends. For
+static objects which cover a known area, the quadtree is by far the most useful.
+For dynamic objects which move a lot the spatial hashing grid is more suitable.
+And lastly, the simple grid is really handy when you quickly want to increase
+the speed of a system without having to spend a lot of time writing the
+optimization strategy, which works really well for applications with a smaller
+amount of objects. A combination of multiple strategies could be used to gain
+the benefit of them, like utilizing the quadtree for static terrain while having
+all dynamic objects be stored in a spatial hashing gird.
+
+Then at what point is it worth implementing a more advanced data structure? Well
+that also depends heavily on your particular use case. If the problem only
+involves checking a handful of objects then the naive solution works just fine.
+As seen in \ref{fig:naive} the naive solution is rarely viable as the object
+count grows. The only real way to know when it is worth to implement a new
+optimization strategy is to measure your current solution and make a decision
+based off of that.
 
@@
 
+\section{Appendix}
+The source code written to get these metrics, alongside the python scripts used
+to generate the graphs, can be found at
+\url{https://github.com/faintforge/spatial-partitioning}.
+
 \end{document}

### 2025-03-13 paper/paper.tex
paper: Final revision.

@@
 \usepackage{float}
-\usepackage{hyperref}
+\usepackage{appendix}
+\usepackage[hidelinks]{hyperref}
+
+\hypersetup{
+    colorlinks=false,
+}
 
@@
     optimize objects in space. It can be found in many applications, from
-    simulating galaxies to video games. Checking collision between objects is an
+    simulating galaxies to video games. Checking collisions between objects is an
     expensive operation that quickly becomes a bottleneck as the number of
@@
 As a consequence of this constraint I had to find a way to optimize this phase.
-That is what sent me down the rabbit hole of spatial partitioning as an
-optimization to do query objects close to each other in space. During my
-research I found out about the three strategies I am comparing in this paper. I
-am doing this paper within the programming course.
+That is what sent me down the rabbit hole of spatial partitioning strategies.
+During my research I discovered the three strategies I am comparing in this
+paper written under the subject of programming.
 
 \section{Purpose \& Scope}
-\label{sec:purpose-and-scope}
 The purpose of this paper is to find the most optimal strategy for dividing up
@@
 
-This study focuses mainly on CPU performance and runtime in exchange for less
-memory efficient implementation. Therefore time based evaluations of insertion,
-query time and general collision testing time will be performed.
+This study focuses mainly on CPU performance and runtime at the cost of higher
+memory usage. Therefore time based evaluations of insertion, query time and
+general collision testing time will be performed.
 
 The strategies are going to be evaluated on time based metrics - insertion,
-query and time spend doing collision testing. These evaluations are going to
-take place with two different distribution strategies of the objects though the
-same seed will be used for all tests so the pseudo-random numbers stay the same.
+query and time spent on collision testing. Two object distributions are going
+to be evaluated. A pseudo-random number generator will be used but given the
+same seed of 1234 so all the pseudo-random numbers stay the same during every
+measurement.
 
 These strategies could be extended into the third dimension but this study
-focuses solely focused on two dimensional space.
+focuses solely on two dimensional space.
 
@@
 \subsection{Spatial Partitioning}
-Spatial partitioning is an optimization technique used to make programs run
-faster. It is often found within physics applications. The theory behind this
-technique is to reduce the number of times one object is checked against
-another. Suppose we have one hundred objects in our world. The naive way to
-check each one of these against the others results in a ten thousand checks.
-Since checks are expensive to make, we want to perform as few of them as
-possible.
+Spatial partitioning is an optimization technique used to make spatial queries
+run faster. It is often found within physics applications. The theory behind
+this technique is to reduce the number of object pairs being checked against
+one another. Suppose we have one hundred objects in our world. The naive way to
+make sure non of them are colliding is to check every object against every other
+object in the world. This results in ten thousand collision checks. Since checks
+are expensive, we want to perform as few of them as possible.
 
 \subsection{Quadtree}
-A quadtree is a recursive data structure, meaning it contains pointers to each
-objects of the same type as itself. It is in the family of tree data structures.
-The data structure is comprised of nodes. Each node has four pointers to other
-nodes, could also be referred to as branches or leaves. This strategy works by
-inserting points into a leaf until the leaf reaches its capacity, then it splits
-into four equally large sub leaves. This subdivision pattern is shown in figure
+A quadtree is a recursive data structure, meaning it contains one or more
+pointers, or references, to an object of the same type as one of its members. It
+is in the family of tree data structures. The data structure is comprised of
+nodes. Each node has four pointers to other nodes, which could also be referred
+to as branches or leaves. This strategy works by inserting points into a leaf
+until the leaf reaches max capacity, then it splits into four equally large sub
+leaves. This subdivision pattern is shown in figure
 \ref{fig:quadtree}.
@@
 \subsection{The Stack and Heap}
-A program uses two different types of memory, the stack and the heap. When the
-program starts a fixed amount of bytes are allocated for the program to use,
-this is what is called the stack. When the program creates a variable it is
-allocated on the stack, which is done by incrementing the stack pointer by the
-size of the variable data type. Since a stack allocation is only a single
-instruction it is incredibly fast. The heap is much larger than the stack but
-slower because you have to request space within it from the operating system.
-Allocating is considered to be a slow operation since there is a lot of overhead
-introduced.
+A program uses two different types of memory, the stack and the heap. When a
+program starts a fixed amount of bytes are allocated for it to use, this is what
+is called the stack. When the program creates a variable it is allocated on the
+stack, which is done by incrementing the stack pointer by the size of the
+variable data type. Since a stack allocation is only a single instruction it is
+incredibly fast. The heap is much larger than the stack but slower because you
+have to request space within it from the operating system. Allocating is
+considered to be a slow operation since there is a lot of overhead introduced.
 
@@
 The static grid is built using a simple array. Indexing into the array is an
-$O(1)$ operation. The caveat with this approach is you are forced to know how
+$O(1)$ operation. The caveat with this approach is you are forced to specify how
 big the area you want the to partition to be before compiling the program. Both
@@
 \subsection{Hash Map}
-A hash map is a data structure which key value pairs allowing for aliasing of
-values. For example, a hash map using strings as keys and integers as values
-could map the key "life" to the value 42, meaning a query into the hash map for
-"life" would return 42. The way this works is by using a cryptographic hashing
-function to generate a unique hash for the object. This hash is a simple number
-which is usually very large. To make the hash useful you run modulo on it to
-make sure it's between zero and the number of buckets inside the hash map. The
-same method is used for retrieval, first running the key through the same
-hashing function, then wrapping it using modulo and returning what lies in the
-bucket. The only caveat with this is there could be collisions within the hash
-map. Two hashes could wrap into the same index. There are a multitude of ways to
-handle collisions like open addressing and separate chaining.
+A hash map is a data structure which stores key-value pairs allowing for
+aliasing of values. For example, a hash map using strings as keys and integers
+as values could map the key "life" to the value 42, meaning a query into the
+hash map for "life" would return 42. The way this works is by using a
+cryptographic hashing function to generate a unique hash for the object. This
+hash is a simple number which is usually very large. To make the hash useful you
+run modulo on it to make sure it's between zero and the number of buckets inside
+the hash map. The same method is used for retrieval, first running the key
+through the same hashing function, then wrapping it using modulo and returning
+what lies in the bucket. The only caveat with this is there could be collisions
+within the hash map. Two hashes could wrap into the same index. There are a
+multitude of ways to handle collisions like open addressing and separate
+chaining.
 
@@
 the object, then run a modulo operation on it giving me a number within the
-array bounds, and then it is inserted into the cell. No collision resolution is
+array bounds, then insert it into the cell. No collision resolution is
 taking place inside the structure meaning two cells could have the same bucket
-index which means queries for the one of the cells returns the objects in the
-other as well.
+index which means queries for one of the cells returns the objects in the other
+as well.
 
@@
 source code. It aids in compiling and linking the program making cross
-compilation and dependency management trivial.
+compilation and dependency management trivial. Since this is a simple project I
+could have used a simple Makefile or even a shell script file but I opted for
+CMake since it's industry standard. This, ultimately, had no effect on the
+programs performance.
 
@@
 Since memory is managed manually in C I could be sure the results would not be
-skewed by a garbage collector running in the middle of a benchmark.
+skewed by a garbage collector running in the middle of a benchmark. Its impact
+is minimal and its fast build times allowed for quickly iterating during
+development.
 
@@
 looking at the code. It also gives a nice visual representation of the program.
+I used SDL2 instead of other similar libraries, such as Raylib, because I'm
+familiar with it. Since the program is dealing with problems concerning two
+dimensional space I found it easier to debug visually which is why SDL2 is a
+dependency. This doesn't affect the results since it was only used during the
+debugging phase of development.
 
 \subsection{Python}
-Python is an interpreted programming language which was used for making a script
-to generate graphs. All the graphs used in this paper was generated from data
+Python is an interpreted programming language which was used in a script to
+generate graphs. All the graphs used in this paper was generated from data
 collected from a C program, which dumped said data into a JSON file, which was
@@
 looking graphs.
-
-\newpage
 
@@
 structure making it ready for rebuilding. Each metric was measured with a range
-of object count, labeled as 'objects' in the graphs, from ten to 2500 with an
-increase of ten objects between each measurement. At each stage was run 32 times
+of object counts, labeled as 'objects' in the graphs, from 10 to 2500 with
+increments of 10 objects between each measurement. At each stage was run 32 times
 and the results were averaged to give smoother graphs. Two benchmarks was run,
@@
 \section{Discussion}
+
+\subsection{Design Choices}
 The design choices for this project was entirely focused on CPU performance
@@
 enough size and pass that as a parameter for the function to modify. Though this
-would work it's not really a good solution, in my subjective opinion. Because
+would work it's not really a good solution, in my subjective opinion, because
 there is guarantee of how many objects could be returned. The better solution
@@
 by the C standard library. The other notable exception is the benchmarking tool
-I made for this. It uses a hash map to store how many times a function was run,
-the total time it ran and the average runtime per call. Because of my underlying
-implementation of the hash map it will require memory allocation when enough
-entries has been added. Not only that, but it was a nested structure which
-created a new hash map for each benchmark scope.
+I developed for this. It uses a hash map to store how many times a function was
+run, the total time it ran and the average runtime per call. Because of my
+underlying implementation of the hash map it will require memory allocation when
+enough entries have been added. Not only that, but it was a nested structure
+which created a new hash map for each benchmark scope.
 
+\subsubsection{Spatial Hashing Grid}
+No collision resolution was implemented for the spatial hashing grid because
+there were very few collisions happening during testing so I didn't pay it much
+mind. I thought the collision handling code would be more expensive than the
+actual performance penalty from not handling the collisions. This was a naive
+way of thinking. I should have implemented and measured instead of making such
+bold assumptions.
+
+\subsection{Compilation}
+One thing I noticed after having generated all the graphs was that the program
+was compiled in debug mode, meaning no compiler optimizations were enabled. If I
+had changed the 'CMAKE\_BUILD\_TYPE' variable in my CMake build script from
+'Debug' to 'Release' the results could be different due to the compiler doing
+clever optimizations to speed up certain things. This could have lead to one
+strategy out performing another. At the time of discovering this, it was too
+late to explore this possibility further.
+
+\subsection{Benchmarking}
 The benchmarks are a thing I would have done differently if I would be redoing
@@
 
+\subsection{Pitfalls of Certain Strategies}
 I found that although quadtrees outperform all the other strategies in this
@@
 are other strategies more suitable for this. The spatial hashing grid solves
-this issue at the cost of performance. Grids come with the same world size
-restriction that the quadtree does but has the benefit of being extremely
-trivial to implement. At with almost all questions the answer to 'In what use
-case are the different algorithms most useful?' the answer is: it depends. For
-static objects which cover a known area, the quadtree is by far the most useful.
-For dynamic objects which move a lot the spatial hashing grid is more suitable.
-And lastly, the simple grid is really handy when you quickly want to increase
-the speed of a system without having to spend a lot of time writing the
-optimization strategy, which works really well for applications with a smaller
-amount of objects. A combination of multiple strategies could be used to gain
-the benefit of them, like utilizing the quadtree for static terrain while having
-all dynamic objects be stored in a spatial hashing gird.
+this issue at the cost of performance. Grids come with the same world size restriction that the quadtree does but has
+the benefit of being extremely trivial to implement.
+
+\subsection{Use Cases}
+As with almost all questions the answer to 'In what use case are the different
+algorithms most useful?' the answer is: it depends. For static objects which
+cover a known area, the quadtree is by far the most useful. For dynamic objects
+which move a lot the spatial hashing grid is more suitable. And lastly, the
+simple grid is really handy when you quickly want to increase the speed of a
+system without having to spend a lot of time writing the optimization strategy,
+which works really well for applications with a smaller amount of objects. A
+combination of multiple strategies could be used to gain the benefit of them,
+like utilizing the quadtree for static terrain while having all dynamic objects
+be stored in a spatial hashing gird.
 
@@
 
-\section{Appendix}
-The source code written to get these metrics, alongside the python scripts used
-to generate the graphs, can be found at
-\url{https://github.com/faintforge/spatial-partitioning}.
+Stack Overflow (2012). What integer hash function are good that accepts an
+integer hash key? Retrieved 2024-12-23, from
+\url{https://stackoverflow.com/questions/664014/what-integer-hash-function-are-good-that-accepts-an-integer-hash-key}
+
+\appendix
+
+\section{Source Code}
+The source code is shipped in a tarball with GZip compression name
+'source.tar.gz'. Once unpacked all source code needed to compile and run it is
+provided. If the SDL2 library isn't installed on the system then the code will
+not compile.
+
+To run the python script to generate the graphs python3 needs to be installed
+alongside the matplotlib library.
+
+A copy of the source code along side some other files can be found on the web
+at \url{https://github.com/faintforge/spatial-partitioning}.
 
